Reproducible and programmatic generation of neuroimaging visualisations

### 2022-04-01 manuscript.Rmd
manuscript update and fig compile

---
title: "Reproducible and programmatic generation of neuroimaging visualisations"
author: "Sidhant Chopra, Löic Labache, Winnie Orchard, [...]"
date: "09/03/2022"
output:
  pdf_document: default
  word_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
library(rgl)
rgl::setupKnitr()
rgl::rgl.open(); rgl::rgl.close() # first window in doc. may not render properly
```

@@

Visualising neuroimaging data is one of the primary ways in which we
evaluate data quality, interpret results and communicate findings. These
visualisations are commonly produced using graphic interface-based (GUI)
tools where individual images are opened and display settings are
manually changed until the desired output is reached. While the choice
to use GUI-based software's has been driven by a perception of
convenience, flexibility and accessibility, there are now many
code-based software packages which are well-documented and don't often
require high-level knowledge of programming. These tools are flexible
and allow for generation of reproducible, high-quality and publication
ready brain visualisations in only a few lines of code (Figure 1),
especially within the R, Python and Matlab environments.

This article will cover three major advantages of using code-generated
visualisations over GUI based tools: replicability, flexibility and
integration. First, by generating figures using code, you increase the
replicability of your figures for yourself, your collaborators and your
readers. Second, code-based software provide more precise controls over
visualisation settings, while also benefiting from advantages of
programming, such as being able to iteratively generate multiple
figures. Finally, being able to integrate figure generation into your
analysis scripts reduces chances of errors, with more advanced tools now
allowing for seamless weaving together of prose and code. Here, we
review the advantages of learning and using programmatic neuroimaging
visualisations, focusing on benefits to replicability, flexibility and
interactivity. We conclude by introducing examples of R-packages which
allow for visualisation of region-of-interest (ROI), voxel, vertex and
edge level data, followed by a brief discussion of limitations and
functionality gaps in code-based brain visualisation tools.

@@

In recent years, there have been multiple large-scale efforts
empirically demonstrating the lack of reproducibility of findings from
neuroimaging data (cite). One common solution proposed for achieving
robust and reliable discoveries has been to encourage scientific output
which can be transparently evaluated and potentially replicated. In
practise, what this usually means is openly sharing detailed methods,
materials, code and data. While there is a trend towards increasing
transparency and code sharing in neuroimaging research (cite), the
sharing of code used to generate figures which include brain renderings
and spatial maps has been relatively neglected. This gap in
reproducibility is partly driven by the fact that brain figures are
often created using a manual process of tinkering with sliders, buttons
and overlays on a GUI, concluding with a screenshot and sometimes
beatification in image processing software like Illustrator. This
process inherently makes neuroimaging visualisations difficult, if not
impossible to replicate, even by the authors who created them.

@@

Being able to exactly replicate your figures via code has serious
advantages beyond good open science practises. In particular, the
ability to reprogram inputs (such as statistical maps) and settings
(such as colour schemes and visual orientations) can streamline your
entire scientific workflow. Being able to code the inputs and settings
allows for the easy production of multiple figures, such as those
resulting from multiple analyses which require similar visualisations.
Likewise, an arduous request from a reviewer to alter the data
processing or analysis becomes less of a burden when the figure can be
re-generated with a few lines of code, as opposed to re-pasting and
re-illustrating them. Having a code-base with reprogrammable inputs can
mean that the generation of visualisations requires less time, energy
and effort than manual GUI-based generation. Keep in mind that the gains
of writing code for your figures are cumulative, and in addition to
improving your programming, you start to build a codebase for figure
generation that you can continue to reuse and share through your
scientific career.

While brain visualisations are often thought of as the end results of
analyses, they also form a vital part of quality control of imaging
data. While tools to automatically detect artefacts, denoise the data
and generate derivatives are becoming more robust, we are not yet at the
stage where visualising the data during processing is no longer need.
Nonetheless, when working with large datasets such as Human Connectome
Project (cite) or UK BioBank (cite), it is infeasible to use traditional
GUI based tool to visually check the data. Knowing how to
programmatically generate brain visualisations can allow you to iterate
your code over each image of a large datasets making quality controls of
each data processing or analysis step accessible and achievable.

@@

Often programming languages such as R, Python and Matlab are used for
the analysis in neuroimaging studies, but the brain visualisation
resulting from these analyses is outsourced to a septate GIU-based
visualisation tools such as FSLeyes, Freeview or ITK-snap. Switching
from analysis to a GUI-based visualisation process be a cumbersome
deviation from the scientific workflow and can make debugging errors
more difficult, as you have to regularly switch program to visually
examine the results of any modifications to analyses. Using the brain
visualisation tools that already exist within your chosen programming
environment can provide instant visual feedback on the impact of
modifications to processing or analyses. While some GUI-based tools do
offer command-line access to generate visualisations, they can lack
flexibility to easily generate publication ready figures and can lose
the benefits provided by your preferred programming environment.

Increasingly popular tools such as R-Markdown, Quarto and Jupyter
notebooks allow for the mixing of prose and code in a single script,
resulting in fully reproducible and publication ready papers. By using
code-based tools available within your preferred environment, brain
visualisations can be directly integrated and embedded within a paper or
report. Journals that publish neuroimaging studies are increasingly
moving towards allowing reproducible manuscripts, including reproducible
figures (cite e-life,f1000,plos), with with some even allowing on-demand
re-running of code using cloud services (cite).

Neuroimaging data are often spatially 3D and can have multiple time
points, adding a 4th dimensional (e.g. fMRI data). Thus, communicating
findings or evaluating quality using static 2D slices is challenging.
While well-curated 3D renderings can help with spatial localisation
(cite madan), in the end, static images can only provide an incomplete
representation of the data. A added advantage of some of code-based
tools is that you can generate 'rich' media like interactive figures or
animations, which can improve scientific communication of findings. Many
code-based software tools now allow for the generation of animated brain
visualisations (cite blog) or interactive widgets which allow users to
zoom, rotate and scroll through slices. Linking to or even embedding
these videos or interactive figures in papers can greatly enhance the
communication of findings.

@@

##### *Voxel-level visualisations.*

The `oro.nifti` package allows for the reading, writing, manipulation
and visualisation of voxel-level imaging data of either *nifti*,
*analyze* or *afni* formats. In particular, the `image`, `slice` and
`orthographic` functions allow users to display the desired number of
slices in the desired orientations, while also providing precise control
over overlaid images, colour scales, legend placement and other
aesthetic settings (Fig 1A). Smoothing and re-sampling of images is
sometimes required for visualisation purposes, which can be achieved via
the *`ANTsR`* /`extrantsr` or `fslr` packages.

@@

##### *Vertex-level visualisations.*

@@

##### *ROI-level visualisations.*

@@

@@

```{r Figure1 }
library(patchwork)
library(ggplot2)

fig1a <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1a.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1b <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1b.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1c <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1c.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1d <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1d.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1e <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1e.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1f <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1f.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1g1 <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1g1.png"),
                                                        width=ggplot2::unit(1,"npc"),
                                                        height=ggplot2::unit(1,"npc")))
fig1g2 <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1g2.png"),
                                                        width=ggplot2::unit(1,"npc"),
                                                        height=ggplot2::unit(1,"npc")))

fig1h <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1h.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

fig1i <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG("data/fig1i.png"),
                                                       width=ggplot2::unit(1,"npc"),
                                                       height=ggplot2::unit(1,"npc")))

patchwork <- fig1a/fig1b|(fig1c|fig1f)/fig1d|(fig1e|fig1g2)/fig1g1|fig1h/fig1i

@@

@@

vis.subject.morph.native(subjects_dir = subjects_dir, 
                         subject_id = 'subject1', 
                         measure = 'thickness', 
                         hemi='both', 
                         views = 't4',
                         draw_colorbar = "horizontal",
                         cortex_only = T, surface = "white", 
                         makecmap_options = list('colFn'=colourmap))
rgl.snapshot("data/fig1d.png")
rgl.close()

>>>>

---
title: "Reproducible and programmatic generation of neuroimaging visualisations"
author: "Sidhant Chopra, Löic Labache, Winnie Orchard, [...]"
date: "09/03/2022"
output:
  html_document: default
  word_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)
```

@@

Visualising neuroimaging data is one of the primary ways in which we
evaluate data quality, interpret results and communicate findings. These
visualisations are commonly produced using graphic interface-based (GUI)
tools where individual images are opened and display settings are
manually changed until the desired output is reached. The choice to use
GUI-based software's has been driven by a perception of convenience,
flexibility and accessibility. However now there are also many
code-based software packages which are well-documented and don't often
require high-level knowledge of programming, making them more accessible
to the neuroimaging community. These tools are flexible and allow for
the generation of reproducible, high-quality and publication-ready brain
visualisations in only a few lines of code (Figure 1), especially within
the R, Python and Matlab environments.

This article will cover three major advantages of using code-generated
visualisations over GUI based tools: replicability, flexibility and
integration. First, by generating figures using code, you increase the
replicability of your figures for yourself, your collaborators and your
readers. Second, code-based software provides more precise controls over
visualisation settings, while also benefiting from advantages of
programming, such as being able to iteratively generate multiple
figures. Finally, being able to integrate figure generation into your
analysis scripts reduces chances of errors, with more advanced tools now
allowing for seamless weaving together of prose and code. Here, we
review the advantages of learning and using programmatic neuroimaging
visualisations, focusing on benefits to replicability, flexibility and
intergration. We conclude by introducing examples of R-packages which
allow for visualisation of statistics at region-of-interest (ROI),
voxel, vertex and edge level data, followed by a brief discussion of
limitations and functionality gaps in code-based brain visualisation
tools.

@@

In recent years, there have been multiple large-scale efforts
empirically demonstrating the lack of reproducibility of findings from
neuroimaging data (cite poldrack). One common solution proposed for
achieving robust and reliable discoveries has been to encourage
scientific output which can be transparently evaluated and independently
replicated. In practise, what this usually means is openly sharing
detailed methods, materials, code and data. While there is a trend
towards increasing transparency and code sharing in neuroimaging
research (cite), the sharing of code used to generate figures which
include brain renderings and spatial maps has been relatively neglected.
This gap in reproducibility is partly driven by the fact that brain
figures are often created using a manual process of tinkering with
sliders, buttons and overlays on a GUI, concluding with a screenshot and
sometimes beatification in image processing software like Illustrator.
This process inherently makes neuroimaging visualisations difficult, if
not impossible to replicate, even by the authors who created them.

@@

Being able to exactly replicate your figures via code has serious
advantages beyond good open science practises. In particular, the
ability to reprogram inputs (such as statistical maps) and settings
(such as colour schemes and visual orientations) can streamline your
entire scientific workflow. Changing inputs and settings via code allows
for the easy production of multiple figures, such as those resulting
from multiple analyses which require similar visualisations. While some
GUI-based tools do offer command-line access to generate visualisations,
they can lack flexibility to easily generate publication ready figures
and can lose the benefits, such as iteration, provided by your preferred
programming environment. A simple for-loop or copying and pasting the
code with the input and/or setting-of-interest modified can be a
powerful method for exploring visualisation options or creating
multi-panel figures. Likewise, an arduous request from a reviewer to
alter the data processing or analysis becomes less of a burden when the
figure can be re-generated with a few lines of code, as opposed to
re-pasting and re-illustrating them. Having a code-base with
reprogrammable inputs can mean that the generation of visualisations
requires less time, energy and effort than manual GUI-based generation.
This also makes it easier to make figures for each subsequent project
that you work on. Keep in mind that the gains of writing code for your
figures are cumulative, and in addition to improving your programming,
you start to build a code-base for figure generation that you can
continue to reuse and share through your scientific career.

Precise controls via code over visualization settings, such as colour
schemes, legend placement and camera angles, can provide you with much
greater flexibility over visualisations. Nonetheless, part of the appeal
of GUI-based tools is that the pre-sets for such settings can provide a
useful starting point and reduce the decision burden on the user.
However, similar pre-sets are often available in the from of default
settings in most code-based packages, negating the need for the user to
manually enter each and every choice of setting required for making the
image. Most code-based tools also come with detialed documentation, or
even entire papers (cite ggseg and ciftitools) which provide examples of
figures which can be used as starting points or templates for new users.
As the popularity of code-sharing for figure increases, there will be an
cornucopia of templates that can be used as starting points for new
figures.

While brain visualisations are often thought of as the end results of
analyses, they also form a vital part of quality control of imaging
data. Tools to automatically detect artefacts, de-noise the data and
generate derivatives are becoming more robust, but we are not yet at the
stage where visualising the data during processing is no longer need.
Nonetheless, when working with large datasets such as Human Connectome
Project (cite) or UK BioBank (cite), it is unfeasible to use traditional
GUI-based tool to visually check the data. The time it takes to open a
single file and achieve the desired visualisation settings vastly
compounds when working with large datasets. Knowing how to
programmatically generate brain visualisations can allow you to iterate
your visualisation code over each image of a large datasets making
checking the quality of each data processing or analysis step accessible
and achievable. The visual outputs of each iteration can be complied
into accessible documents which can be easily scrolled, with more
advanced usage allowing for the creation of interactive HTML reports,
similar to those created by tools like mriQC (cite). Increasing capacity
to conduct visual quality control on larger datasets can increase the
identification of processing errors and result in more reliable and
valid findings from your data.

@@

Often programming languages such as R, Python and Matlab are used for
the analysis in neuroimaging studies, but the brain visualisation
resulting from these analyses is outsourced to separate GIU-based
visualisation tools such as FSLeyes, Freeview or ITK-snap. Switching
from your analysis environment to a GUI-based visualisation process be a
cumbersome deviation from the scientific workflow. This can make
debugging errors more difficult, as you have to regularly switch program
to visually examine the results of any modifications to analyses. Using
the brain visualisation tools that already exist within your chosen
programming environment can provide instant visual feedback on the
impact of modifications to processing or analyses.

Increasingly popular tools such as R-Markdown, Quarto and Jupyter
Notebook allow for the mixing of prose and code in a single script,
resulting in fully reproducible and publication ready papers. By using
code-based tools available within your preferred environment, brain
visualisations can be directly integrated and embedded within a paper or
report. Some journals that publish neuroimaging studies are moving
towards allowing reproducible manuscripts, including reproducible
figures (cite e-life,f1000,plos), with some even allowing on-demand
re-running of code using cloud services (cite).

Neuroimaging data are often spatially 3D and can have multiple time
points, adding a 4th dimension (e.g. fMRI data). Thus, communicating
findings or evaluating quality using static 2D slices is challenging.
While well curated 3D renderings can help with spatial localisation
(cite madan), in the end, static 2D images can only provide an
incomplete representation of 3D or 4D data. A added advantage of some of
code-based tools is that you can generate 'rich' media like interactive
figures or animations, which can improve scientific communication of
findings. Many code-based software tools now allow for the generation of
animated brain visualisations (cite blog) or interactive widgets which
allow users to zoom, rotate and scroll through slices. Linking to or
even embedding these videos or interactive figures in papers can greatly
enhance the communication of findings.

Neuroimaging data are often spatially 3D and can have multiple time
points, adding a 4th dimension (e.g. fMRI data). Thus, communicating
findings or evaluating quality using static 2D slices is challenging,
and may not be the best representation of the data, or the
interpretations. While well-curated 3D renderings can help with spatial
localisation (see madan), in the end, static images can only provide an
incomplete representation of the data, and forces researchers to choose
the best angle to show, which often involves compromising one result for
another. An added advantage of some of the code-based tools is that you
can generate 'rich' media like interactive widgets -- figures or
animations, which allow users to zoom, rotate and scroll through slices.
Interacting with a figure in this way can improve scientific
communication of findings. Linking to or even embedding these videos or
interactive figures in papers can greatly enhance the communication of
findings and make your paper more engaging for the reader. Such rich
brain visualisations lend themselves to being embedded or shared on
science communication mediums other than academic papers - such as
presentations, web-sites and social media - all of which can promote the
communication of your research with peers and a reach larger audiences.

@@

The following four sections provide brief examples packages and
functions available in R for visualising voxel, vertex, ROI and
edge-level data. A more didactic introduction to these tools is provided
in a [collection of beginner friendly
R-Notebooks](https://github.com/sidchop/neuRo-vis-BrainHack-2021).

```{r Figure1 , echo=FALSE}
library(patchwork)
library(ggplot2)

#loop this?
fig_name <- c('fig1a', 'fig1b', 'fig1c', 'fig1d', 'fig1e', 'fig1f', 'fig1g1', 'fig1g2', 'fig1h', 'fig1i')
figs <- list()

for (f in 1:length(fig_name)) { 
  figs[[f]] <- ggplot() + annotation_custom(grid::rasterGrob(png::readPNG(paste0('data/',fig_name[f],'.png')),
                                                             width=ggplot2::unit(1,"npc"),
                                                             height=ggplot2::unit(1,"npc")))
}

patchwork <- figs[[1]]/figs[[2]]|(figs[[3]]|figs[[6]])/figs[[4]]|(figs[[5]]|figs[[8]])/figs[[7]]|figs[[9]]/figs[[10]]

#& theme(text = element_text('Times New Roman'))

\
*[Figure 1 - Add Caption]*

##### Voxel-level visualisations.

The `oro.nifti` package allows for the reading, writing, manipulation
and visualisation of voxel-level imaging data of either nifti, analyze
or afni formats. In particular, the `image`, `slice` and `orthographic`
functions allow users to display the desired number of slices in the
desired orientations, while also providing precise control over overlaid
images, colour scales, legend placement and other aesthetic settings
(Fig 1A). Smoothing and re-sampling of images is sometimes required for
visualisation purposes, which can be achieved via the `ANTsR`
/`extrantsr` or `fslr` packages.

@@

##### Vertex-level visualisations.

@@

##### ROI-level visualisations.

@@

-   Leged placement and creation

@@

@@

#weird angle

@@

vis.subject.morph.native(subjects_dir = subjects_dir, 
                         subject_id = 'subject1', 
                         measure = 'thickness', 
                         hemi='both', 
                         views = 't9',
                         draw_colorbar = "horizontal",
                         cortex_only = T, surface = "white", 
                         makecmap_options = list('colFn'=colourmap))
rgl.snapshot("data/fig1d.png")
rgl.close()
